---
title: "DTP vaccinations"
author: |
  G.J.R. Weydemuller (student number)  
  L. Waaijenberg (2862154)  
  P. Gandra Rodrigues (2870539)  
  Q. Faber (2861921)  
  R.A. Kali (2850680)  
  T. Adam (student number)  
  T.M.V. Verbeek (student number)
date: "`r Sys.Date()`"
output: html_document
---


```{r package_install, include=FALSE}
options(repos = c(CRAN = "https://cloud.r-project.org"))
install.packages("tidyverse")
install.packages("readxl")
install.packages("dplyr")
install.packages("sf")
install.packages("rnaturalearth")
install.packages("rnaturalearthdata")
install.packages("ggplot2")
install.packages("viridis")
install.packages("grid")
```

```{r setup, include=FALSE}
library(tidyverse)
library(readxl)
library(dplyr)
library(sf)
library(rnaturalearth)
library(rnaturalearthdata)
library(ggplot2)
library(viridis)
library(grid)
```


# Report

**G.J.R. Weydemuller** (student number)  
**L. Waaijenberg** (2862154)  
**P. Gandra Rodrigues** (2870539)  
**Q. Faber** (2861921)  
**R.A. Kali** (2850680)  
**T. Adam** (student number)  
**T.M.V. Verbeek** (student number)  
---
**Tutor:** J.F. Fitzgerald  
**Tutorial Group:** Tutorial group 1  
**Course:** Programming for Economists  
**Institute:** Vrije Universiteit Amsterdam



# Problem Motivation and Literature

## Why is DTP vaccination a social problem?

DTP vaccination, protecting against diphtheria, tetanus, and pertussis, is a cornerstone of global public health. Despite its proven effectiveness and inclusion in nearly all national immunization schedules, millions of children worldwide remain under-vaccinated or entirely unvaccinated. This gap in coverage leads to preventable illness, long-term disability, and death, particularly in low- and middle-income countries. The issue is not merely medical but deeply social, as it reflects and reinforces inequalities in access to healthcare, education, and infrastructure.

Vaccination coverage is also a proxy for the strength of a country’s healthcare system. Low DTP coverage often signals broader systemic issues such as poverty, conflict, misinformation, and weak governance. These factors disproportionately affect vulnerable populations, making DTP vaccination a critical indicator of social justice and equity.

## Research Question

**"How does government healthcare spending per capita correlate with national vaccination coverage rates for routine immunizations in low-, middle-, and high-income countries?"**

## Sources Identifying DTP Vaccination as a Social Problem

- The **World Health Organization (WHO)** and **UNICEF** have jointly warned that vaccine-preventable diseases, including diphtheria, are resurging due to declining immunization coverage, misinformation, and funding cuts. They emphasize that millions of children are at risk, and urgent investment in immunization is needed to prevent a reversal of decades of progress (WHO Media Team, 2025).

- **UNICEF** outlines the severe consequences of DTP-preventable diseases. Diphtheria can cause heart and nerve damage, tetanus is often fatal even with treatment, and pertussis can lead to pneumonia and death in infants. These diseases disproportionately affect children in low-resource settings, highlighting the social dimension of under-vaccination (UNICEF, sd).

- A **UNICEF data report** on DTP vaccine dropout rates shows that millions of children start but do not complete the DTP series. This dropout is especially prevalent in low-income countries and is linked to systemic barriers such as healthcare access, education, and infrastructure (Wang et al., 2019).

## What Has Not Been Fully Researched

While many studies have examined DTP coverage and its determinants, there is limited comparative analysis across income groups that links government healthcare spending per capita directly to national DTP vaccination rates. Most existing research focuses on either micro-level determinants (e.g., parental education, rural access) or macro-level trends without integrating economic policy variables.

## Contribution of This Report

This report aims to fill that gap by analyzing how government healthcare spending correlates with DTP vaccination coverage across low-, middle-, and high-income countries. This research will provide new insights into whether increased public investment in health translates into better immunization outcomes. This could inform both global health policy and economic development strategies, especially in resource-constrained settings.

# Data Sourcing

## Load in Data

```{r}

url_spending <- "https://www.dropbox.com/scl/fi/01df0ly3ornjgcis9kwsk/CHE-per-region.csv?rlkey=nwvmxv33dc3uvp9pnil4tqimk&st=5zgoazd8&dl=1"

url_coverage <- "https://www.dropbox.com/scl/fi/08rf7e6sj3hcq3e5fnp8l/datasetcoverage.xlsx?rlkey=x9iwcifdi2vmqa3j4oy1zs7p3&st=g5kdhaib&dl=1"

url_childdeath <- "https://www.dropbox.com/scl/fi/wuopfxxyux4dccc9rbibn/child-death.csv?rlkey=mfqw1b76522vtjqxcz7grvtmb&st=v0q288at&dl=1"
```

```{r include=FALSE}
# Download het bestand tijdelijk en lees het in
temp_file <- tempfile(fileext = ".xlsx")
download.file(url_coverage, destfile = temp_file, mode = "wb")
df <- read_excel(temp_file)

# Read the CSV into a dataframe
raw_healthspending_data <- read.csv(url_spending)
raw_healthspending_data <- raw_healthspending_data %>%
select(Indicator, Location, Period, Value)


# Read the CSV into a dataframe
raw_childdeath_data <- read.csv(url_childdeath)
```


## Provide a short summary of the datasets
```{r echo=FALSE}
head(raw_healthspending_data)
head(df)
```